`ggplot2` part 01
========================================================

```{r include = FALSE}
## sometimes necessary until I can figure out why loaded packages are leaking
## from one file to another, e.g. from block91_latticeGraphics.rmd to this file
if(length(yo <- grep("gplots", search())) > 0) detach(pos = yo)
if(length(yo <- grep("gdata", search())) > 0) detach(pos = yo)
if(length(yo <- grep("gtools", search())) > 0) detach(pos = yo)
```

SO UNDER DEVELOPMENT IT BARELY EXISTS!

> The blind leading the blind? As I've warned you, I'm *much* newer to `ggplot2` than I am to `lattice` or even `plyr`. I am showing you what it looks like to stay current: periodically you have to learn new tools. It's also par for the course that I've waited a ridiculous amount of time to take this plunge. Better late than never.

### Optional getting started advice

*Ignore if you don't need this bit of support.*

This is one in a series of tutorials in which we explore basic data import, exploration and much more using data from the [Gapminder project](http://www.gapminder.org). Now is the time to make sure you are working in the appropriate directory on your computer, perhaps through the use of an [RStudio project](block01_basicsWorkspaceWorkingDirProject.html). To ensure a clean slate, you may wish to clean out your workspace and restart R (both available from the RStudio Session menu, among other methods). Confirm that the new R process has the desired working directory, for example, with the `getwd()` command or by glancing at the top of RStudio's Console pane.

Open a new R script (in RStudio, File > New > R Script). Develop and run your code from there (recommended) or periodicially copy "good" commands from the history. In due course, save this script with a name ending in .r or .R, containing no spaces or other funny stuff, and evoking "ggplot2" and "lattice".

### Load the Gapminder data and `ggplot2` (and `plyr`)

Assuming the data can be found in the current working directory, this works:
```{r, eval=FALSE}
gDat <- read.delim("gapminderDataFiveYear.txt")
```

Plan B (I use here, because of where the source of this tutorial lives):
```{r}
## data import from URL
gdURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gDat <- read.delim(file = gdURL)
```

Basic sanity check that the import has gone well:
```{r}
str(gDat)
```

If you have not already done so, now is the time to install `ggplot2`.

```{r eval = FALSE}
install.packages("ggplot2", dependencies = TRUE)
```

Load the `ggplot2` and `plyr` packages.
```{r}
library(ggplot2)
library(plyr)
```

### JB's early notes to self

[Karthik and others say to NOT use `qplot()`](http://inundata.org/2013/04/10/a-quick-introduction-to-ggplot2/). It just makes for more re-learning when you graduate to `ggplot()`. No training wheels for us then.

Pass your data to `ggplot()` as a data.frame, duh. But wait, it gets better: it's the *only* way to provide data to `ggplot()`. Love that.

"aesthetic attributes, visual properties that affect the way observations are displayed" p. 12

The word "aesthetic" makes me think of colors, line types, and symbol shapes. But remember the mapping of a variable into the horizontal location of a plot element -- so saying `x = thisVar` -- is also considered part of specifying the "aesthetic". Strange but true.

"For every aesthetic attribute, there is a function, called a *scale*, which maps data values to valid values for that aesthetic." p. 12-3.

"... `geom`. Geom, short for geometric object, describes the type of object that is used to display the data." p. 13-4

The storage and use of returned graphical objects is much more comfortable for me in `ggplot2` than it's ever been with `lattice`. The `this + that` syntax makes this a total win.

You set something to a constant. You map something to a variable. If you're setting something, it is likely you'll need to protect it with `I()`.

When things from the book don't work, remember `ggplot2` has evolved since it was written.

  * <http://cloud.github.com/downloads/hadley/ggplot2/guide-col.pdf>
  
"The grammar is also useful because it suggests the high-level aspects of a plot that *can* be changed, giving you a framework to think about graphics, and hopefully __shortening the distance from mind to paper__." p. 27 I like that last phrase



### Prompts for in class work and homework

Scatterplot + smooth curve (yes, a straight line is just a special case).

  * 2.5.1 Adding a smoother to a plot. Confidence band on/off. loess vs GAM vs polynomial vs spline. Wiggly vs smooth. Robust?

Line plot

  * Plot connect-the-dots lines of lifeExp over year for different countries
  
Stripplot + level-specific means or median of the quantitative variable
  
  * Add jitter. Control how wide the jitter it. Play with transparency. How to emulate `lattice`'s `type = "a", fun =`?
  
Densityplot

  * Play with smoothness. Emulate `groups =`, i.e. superpose different densityplots.
  
Histogram

  * Play with bidwidth or break points. (Try to) emulate `groups =`, i.e. superpose different densityplots
  
Bar chart

  * Make a bar chart of the number of countries within each continent.
  * Juxtapose with same from `lattice` which we really have not explored explicitly in this course.

### References

ggplot2: Elegant Graphics for Data Analysis [available via SpringerLink](http://ezproxy.library.ubc.ca/login?url=http://link.springer.com.ezproxy.library.ubc.ca/book/10.1007/978-0-387-98141-3/page/1) by Hadley Wickham, Springer (2009) | [online docs (nice!)](http://docs.ggplot2.org/current/) | [author's website for the book](http://ggplot2.org/book/), including all the code | [author's landing page for the package](http://ggplot2.org)

<div class="footer">
This work is licensed under the  <a href="http://creativecommons.org/licenses/by-nc/3.0/">CC BY-NC 3.0 Creative Commons License</a>.
</div>
