Go back to [STAT545A home](current.html)

Homework #4 Visualize a quantitative variable with `lattice` 
========================================================

> Follow the [existing homework submission instructions](hw00_instructions.html) UP TO THE LINK SUBMISSION PART. By 9:30am Monday September 30, be prepared to share your links via a Google doc.

  * Start with the Gapminder data as provided in [`gapminderDataFiveYear.txt`](http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt).
  * Pick at least two data aggregation tasks from [homework 3](hw03_dataAggregation.html), also listed below. __You decide what to work on__ but spend ~2 hours on this. Of course you are welcome to do more if you wish.
  * Try to pick tasks that you did not do, if you left any undone. Read / modify data aggregation code from a fellow student (best!) or JB.
    - If applicable, state and link to the student's code and report.
    - Comment on their HW 3 report and/or code if you wish. Be kind but constructive.
    - Notice what makes code hard vs. easy to read and re-use.
  * Use the `lattice` package to create "companion" graphics, at least one per data aggregation task.
    - Your figure does not have to depict every last number from the data aggregation result.
    - The figure can illustrate things that are hard/impossible to see in the data aggregation result.
    - Goal: the table and the figure complement each other!
    - Notice which data formats make better inputs for plotting functions.
  * Write up with R Markdown.
    - Include a narrative, written in English prose. Don't just show code and results.
    - Expose your code, i.e. use `echo = FALSE` very sparingly.
  * DUE: Before class begins 9:30am Monday September 30.
    - Compile into an HTML report. Follow the naming convention. Publish the HTML report on the web somewhere, such as on RPubs. Make the slug follow the naming convention.
    - Publish the R Markdown file as a Gist. Follow the naming convention.
    - Share your links by editing the Google doc mentioned above.
 
Tips

  * Perform a superficial check that data import went OK.
  * Give informative, short variable names, so you don't need to fiddle with `xlab =` etc.
  * Attend to the order of your factor levels.
  * Feel free to use `lattice` functions we haven't explicitly covered yet. In particular, `xyplot()` might be handy and basic use is pretty easy. 
  
### Menu of data aggregation tasks.

#### Choose your own adventure.

You are welcome to improvise. Wherever I say "life expectancy", you could substitute "GDP per capita". When I say "max", feel free to look at the 0.9 quantile. You get the idea.

```{r include = FALSE}
## I format my code intentionally!
## do not re-format it for me!
opts_chunk$set(tidy = FALSE)

## toggle to turn solution chunks on/off

## Assign HW: This file would live in private/ and not be in public repo and
## would not compiled by fileherd. To make student-facing assignment, uncomment
## the line below, knit this file, and copy the resulting .md file up one level,
## where fileherd will pick it up and compile to HTML and we can push to web.
## Only the data aggregation task prompts will appear.

## Reveal solutions: Move this file up into main directory. Leave the line below
## commented out. fileherd will compile questions AND solutions normally.

## Reset course for future: Hmmmm.... now the solutions are "out there on the
## web". I probably should have thought of that sooner. Worry about that later.

opts_chunk$set(eval = FALSE, echo = FALSE)
```

<!---
#### Load the Gapminder data and the `plyr` and `xtable` packages
--->
```{r}
## data import from URL
gdURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt"
gDat <- read.delim(file = gdURL)
library(plyr)
library(xtable)
str(gDat)
## define a function for converting and printing to HTML table
htmlPrint <- function(x, ...,
                      digits = 0, include.rownames = FALSE) {
  #if(!hasArg("digits")) digits <- 0
  print(xtable(x, digits = digits, ...), type = 'html',
        include.rownames = include.rownames, ...)
  }
```
<!--- I wrote a function `htmlPrint()` to create an HTML table and print it. I am also trying blow your mind with the power of the `...` argument. That is why I can still pass arguments like `include.rownames = TRUE` whenever I feel like it. Here I also make the digits default to 0, which seemed best for the exercises below.
--->

#### Depict the maximum and minimum of GDP per capita for all continents.

```{r results = 'asis'}
foo <- ddply(gDat, ~ continent, summarize,
             minGdpPercap = min(gdpPercap), maxGdpPercap = max(gdpPercap))
htmlPrint(arrange(foo, minGdpPercap))
```

```{r results = 'asis'}
foo <- ddply(gDat, ~ continent, function(x) {
  gdpPercap <- range(x$gdpPercap)
  return(data.frame(gdpPercap, stat = c("min", "max")))
  })
htmlPrint(foo)
```
#### Look at the spread of GDP per capita within the continents.

```{r results = 'asis'}
foo <- ddply(gDat, ~ continent, summarize,
             sdGdpPercap = sd(gdpPercap), madGdpPercap = mad(gdpPercap),
             iqrGdpPercap = IQR(gdpPercap))
htmlPrint(arrange(foo, sdGdpPercap))
```

#### Compute a trimmed mean of life expectancy for different years.

> It's fine to substitute something more pedestrian, like a median. Or try the trimmed mean thing! In that case, try to handle your reporting of the trim level as professionally as possible.

```{r results = 'asis'}
jTrim <- 0.2
foo <- ddply(gDat, ~ year, summarize, tMean = mean(lifeExp, trim = jTrim))
htmlPrint(arrange(foo, tMean))
```

#### How is life expectancy changing over time on different continents?

```{r results = 'asis'}
htmlPrint(ddply(gDat, ~ continent + year,
                summarize, medLifeExp = median(lifeExp)))
```

```{r results = 'asis'}
foo <- daply(gDat, ~ year + continent,
             summarize, medLifeExp = median(lifeExp))
#str(foo)
foo <- as.data.frame(foo)
#str(foo) ## there's still some goofiness, ie the variables are lists .... hmmmm
htmlPrint(foo, include.rownames = TRUE)
```

```{r results = 'asis'}
foo <- daply(gDat, ~ continent + year,
             summarize, medLifeExp = median(lifeExp))
htmlPrint(as.data.frame(foo), include.rownames = TRUE)
```

#### Depict the number and/or proportion of countries with low life expectancy over time by continent.

```{r results = 'asis'}
#(bMark <- quantile(gDat$lifeExp, 0.2))
bMark <- 43 # JB wrote this on her 43rd b'day!
htmlPrint(ddply(gDat, ~ continent + year,
                function(x) c(lowLifeExp = sum(x$lifeExp <= bMark))))
```

```{r results = 'asis'}
## note: bMark set in previous chunk
htmlPrint(ddply(gDat, ~ continent + year, function(x) {
  jCount = sum(x$lifeExp <= bMark)
  c(count = jCount, prop = jCount / nrow(x))
}), digits = c(0, 0, 0, 0, 2))
```

#### Find countries with interesting stories.

Find countries with extremely low or high life expectancy in 1952 or that exhibit extremely rapid or slow life expectancy gains. Then plot their data.

Find countries with sudden, substantial departures from the temporal trend in one of the quantitative measures. Then plot their data.

```{r results = 'asis'}
yearMin <- min(gDat$year)
jFun <- function(x) {
  jFit <- lm(lifeExp ~ I(year - yearMin), x)
  jCoef <- coef(jFit)
  names(jCoef) <- NULL
  return(c(intercept = jCoef[1],
           slope = jCoef[2],
           maxResid = max(abs(resid(jFit)))/summary(jFit)$sigma))
  }
maxResids <- ddply(gDat, ~ country + continent, jFun)
foo <- ddply(maxResids, ~ continent, function(x) {
  theMax <- which.max(x$maxResid)
  x[theMax, ]
  })
htmlPrint(foo, digits = c(0, 0, 0, 2, 2, 2))
## I can't resist, I must make a plot
xyplot(lifeExp ~ year | country, gDat,
       subset = country %in% foo$country, type = c("p", "r"))
```

### Student work on HW 3 for ease of access. We'll replace these links w/ HW 4 later.

  * attali-dea [source](https://gist.github.com/daattali/6673361#file-stat545a-2013-hw03_attali-dea-rmd) | [report](http://rpubs.com/daattali/stat545a-2013-hw03_attali-dea) 
  * baik-jon [source](https://gist.github.com/jonnybaik/6667437#file-stat545a-2013-hw03_baik-jon-rmd) | [report](http://rpubs.com/jonnybaik/stat545a-2013-hw03_baik-jon)
  * bolandnazar-moh [source](https://gist.github.com/ArephB/6667983#file-stat545a-2013-hw03_bolandnazar-moh-rmd) | [report](http://rpubs.com/aref/stat545a-2013-hw03_bolandnazar-moh)
  * brueckman-chr EDIT HERE
  * chu-jus [source](https://gist.github.com/JustinChu/6667252#file-stat545a-2013-hw03_chu-jus-rmd) | [report](http://rpubs.com/cjustin/stat545a-2013-hw03_chu-jus)
  * Zachary Daly: 
[source](https://gist.github.com/ZDaly/6666365#file-stat545a-2013-hw03_daly-zac-rmd)
[report](http://rpubs.com/Zdaly/stat545a-2013-hw03_daly-zac)
  * dinsdale-dan [source](https://gist.github.com/danieldinsdale/6665986#file-stat545a-2013-hw03_dinsdale-dan-rmd) | [report](http://rpubs.com/danieldinsdale/stat545a-2013-hw03_dinsdale-dan)
  * gao-wen [source](https://gist.github.com/sibyl229/6668047#file-stat545a-2013-hw03_gao-wen-rmd) | [report](http://rpubs.com/less/stat545a-2013-hw03_gao-wen)
  * Matthew Gingerich: [source](https://gist.github.com/MattGingerich/6667676#file-stat545a-2013-hw03_gingerich-mat-rmd) | [report](http://rpubs.com/majugi/stat545a-2013-hw03_gingerich-mat)
  * hu-yum [source](https://gist.github.com/smilecat/6666754#file-stat545a-2013-hw03_hu-yum-rmd) | [report](http://rpubs.com/smilecat/stat545a-2013-hw03_hu-yum)
  * jewell-sea [source](https://gist.github.com/jewellsean/bf3c28b63e6f99953153#file-stat545a-2013-hw03_jewell-sea-rmd) | [report](http://rpubs.com/jewellsean/stat545a-2013-hw03_jewell-sea)
  * johnston-reb [source](https://gist.github.com/rebjoh/6667299#file-stat545a-2013-hw03_johnston-reb-rmd) | [report](http://rpubs.com/rljohn/stat545a-2013-hw03_johnston-reb)
  * mahdiar khosravi [source](https://gist.github.com/Mahdiark/6671508#file-stat545a-2013-hw03_khosravi-mah-rmd) | [report](http://rpubs.com/mahdiar/stat545a-2013-hw03_khosravi-mah)
  * Wooyong Lee: 
[source](https://gist.github.com/folias/6661622#file-stat545a-2013-hw03_lee-woo) | [report](http://rpubs.com/folias/stat545a-2013-hw03_lee-woo)
  * liao-wei: [source](https://gist.github.com/feiba/6674299#file-stat545a-2013-hw03_liao_wei-rmd) | [report](http://rpubs.com/winson/stat545a-2013-hw03_liao_wei)
  * ma-hui [source](https://gist.github.com/horsehuiting/6673172#file-stat545a-2013-hw03_ma-hui-rmd) | [report](http://rpubs.com/Huiting/stat545a-2013-hw03_ma-hui)
  * meng-viv [source](https://gist.github.com/vmeng321/6667418#file-stat545a-2013-hw03_meng-viv-rmd) | [report](http://rpubs.com/vmeng321/stat545a-2013-hw03_meng-viv)
  * mohd abul basher-abd [source](https://gist.github.com/atante/6666678#file-stat545a-2013-hw03_mohd-abul-basher-abd-rmd) | [report](http://rpubs.com/meitantei/stat545a-2013-hw03_mohdabulbasher-abd) 
  * ni-jac [source](https://gist.github.com/jacknii/6663292#file-stat545a-2013-hw03_ni-jac-rmd) | [report](http://rpubs.com/jackni/stat545a-2013-hw03_ni-jac)
  * Christian Okkels [source](https://gist.github.com/cbokkels/6654193#file-stat545a-2013-hw03_okkels-chr-rmd) | [report](http://rpubs.com/cbokkels/stat545a-2013-hw03_okkels-chr)
  * Greg Owens: [source](https://gist.github.com/opsin/6666283#file-stat545a-2013-hw03_owens-greg-rmd) | [report](http://rpubs.com/opsin/stat545a-2013-hw03_owens-greg)
  * Mina Park: [source](https://gist.github.com/parkm87/6665438#file-stat545a-2013-hw03_park-min-rmd) | [report](http://rpubs.com/parkm87/stat545a-2013-hw03_park-min)
  * spencer-nei [source](https://gist.github.com/neilspencer/6558151#file-stat545a-2013-hw03_spencer-neil-rmd) | [report](http://rpubs.com/neil_spencer/stat545a-2013-hw03_spencer-nei)
  * wang-ton [source](https://gist.github.com/yzhxh/6670081#file-stat545a-2013-hw03_wang-ton-rmd) | 
[report](http://rpubs.com/yzhxh/stat545a-2013-hw03_wang-ton)
  * Leah Weber: [source](https://gist.github.com/lweber21/6667096#file-stat545a-2013-hw03_weber-lea-rmd) | [report](http://rpubs.com/lweber21/stat545a-2013-hw03_weber-lea)
  * woollard-geo [source](https://gist.github.com/geoffwoollard/6666457#file-stat545a-2013-hw03_woollard-geo-rmd) | [report](http://rpubs.com/gwoollard/stat545a-2013-hw03_woollard-geo)
  * xue-xinxin [source](https://gist.github.com/xxue/6663673#file-stat545a-2013-hw03_xue-xinxin-rmd) | [report](http://rpubs.com/xxue/8829) 
  * yuen-ama [source](https://gist.github.com/amandammor/6666007#file-stat545a-2013-hw03_yuen-ama-rmd) | [report](http://rpubs.com/amandammor/stat545a-2013-hw03_yuen-ama)
  * zhang-yim [source](https://gist.github.com/zym268/6667601#file-stat545a-2013-hw03_zhang-yim-rmd)| [report](http://rpubs.com/zym268/STAT545A-2013-hw03_zhang-yim)
  * zhang-jon [source](https://gist.github.com/jzhang722/6665486#file-stat545a-2013-hw03_zhang-jon-rmd) | [report](http://rpubs.com/jzhang722/stat545a-2013-hw03_zhang-jon)
  * zhang-yif 
[source](https://gist.github.com/dora7870/6630819#file-stat545a-2013-hw03_zhang-yif-rmd) | [report](http://rpubs.com/dora7870/stat545a-2013-hw03_zhang-yif)
  * zhang-jin [source](https://gist.github.com/0527zhangjinyuan/6673340#file-stat545a-2013-hw03_zhang-jin-rmd) | [report](http://rpubs.com/zhangjinyuan/stat545a-2013-hw03_zhang-jin)
  * inskip-jes [source](https://gist.github.com/jinskip/6665098#file-stat545a-2013-hw03_inskip-jes-rmd) | [report](http://rpubs.com/jinskip/stat545a-2013-hw03_inskip-jes)
  * liu-yan [source](https://gist.github.com/swallow0001/6673645#file-stat545a-2013-hw3_liu) | [report](http://rpubs.com/swallow0001/STAT545a-2013-hw3_Liu)
  * haraty-mon EDIT HERE
  * yuen-mac [source](https://gist.github.com/myuen/6672095#file-stat545a-2013-hw03_yuen-mac-rmd) | [report](http://rpubs.com/myuen/stat545a-2013-hw03_yuen-mac)
  

<div class="footer">
This work is licensed under the  <a href="http://creativecommons.org/licenses/by-nc/3.0/">CC BY-NC 3.0 Creative Commons License</a>.
</div>